Brief Announcement: An Improved Lower Bound for Dynamic Balanced Graph Partitioning

### 2020-02-17 paper/paper.tex
partition VS partitioning

  In the online Graph RePartitioning problem, we are given a sequence of pairwise communication requests between n nodes, revealed online.
  The objective is to service these requests by partitioning the nodes into $\ell$ clusters, each of size $k$, such that frequently communicating nodes are located in the same cluster.
  The partitioning can be updated by migrating nodes between clusters for a fixed cost.
  The goal is to jointly minimize the amount of inter-cluster communication and migration cost.

  So far, the best known lower bound for the competitive ratio of
	deterministic algorithms was $\Omega(k)$,
 by reduction from online paging.
  We improve this result significantly by providing the lower bound of $\Omega(k\cdot \ell)$.
  We follow up with an optimal algorithm for $k=3$.
  Finally,
  we provide a $O(k\cdot \ell)$-competitive algorithm for the variant
  where the optimal configuration is always a fixed partition
  that incurs zero communication cost through the whole request sequence.
\end{abstract}

@@

Main technical contribution of this paper is the lower bound $\Omega(k\cdot\ell)$ for the competitive ratio of any deterministic algorithm for online Balanced RePartition problem (see Section~\ref{sec:lowerbound}).
The best known lower bound so far was $\Omega(k)$ \cite{repartition-disc}
which involves only two clusters.

@@

In Section~\ref{sec:ppl}, we study a restricted model of Balanced RePartition, called Perfect Partition.
For that, we present a $O(k\cdot \ell)$-competitive algorithm, which is optimal.
We believe that this result is an important building block on the way to an optimal algorithm for the general model.

@@

The online \emph{Balanced RePartitioning} problem (\OBRP{}) is defined as
follows. We are given a set of $n$ nodes,
initially arbitrarily partitioned into $\ell$~clusters,
each of size~$k$.
The input consists of a sequence of pairwise communication requests
$\sigma = (u_1,v_1),$ $(u_2,v_2),$ $(u_3,v_3), \ldots$,
where a pair $(u_t,v_t)$ indicates that nodes $u_t$ and $v_t$ exchange a fixed amount of data.
Nodes in $C \subset V$ are \emph{collocated}
if they reside in the same cluster.
An algorithm serves a communication request between two nodes
either \emph{locally} at cost~0
if they are collocated,
or \emph{remotely} at cost~1
if they are located in different clusters.
We refer to the two types of requests as \emph{internal}
and \emph{external} requests, respectively.
Before serving a request,
an online algorithm may \emph{re-partition},
%(i.e., \emph{reconfigure}).
i.e.,
it may (optionally) move some nodes into clusters different than their current clusters while respecting the capacity of every cluster. 
Only after a re-partition is completed,
the algorithm serves the  request.
The cost of migrating a node from one cluster to another
is~$\alpha \in \mathbb{Z}^+$.
For any algorithm $ALG$,
its cost,
denoted by $ALG(\sigma)$,
is the total cost of communications and
the cost of migrations performed by $ALG$ while serving the sequence $\sigma$.

@@

%We define the algorithm \TAlg in the following way.
Now we describe the algorithm \TAlg.
It partitions nodes into components, and
intially each node belongs to its own singleton component.
For each pair of nodes, \TAlg maintains a counter, initiated to $0$. 
Upon receiving a request to a pair that is not collocated at one cluster, it increases their counter by $1$.
If the counter for a pair $(u,v)$ reaches $\alpha$, \TAlg merges the components of $u$ and $v$, and moves to the closest component-respecting partition.
If no such partition exists, \TAlg resets all components to singletons, resets all counters to $0$, and ends the phase.
\maciek{Serving remotely: say explicitly}

@@

\begin{proof}
  Observe that when the repartition is triggered by \TAlg, the resulting partition is component-respecting.
  Otherwise, if it does not exist, \TAlg simply ends the phase and performs no repartition.

@@

    Consider a request between $u$ and $v$ that triggered the repartition and let $U$ and $V$ be their respective clusters.
    Note that $U\neq V$,
	 as otherwise the request would not trigger a~re-partitioning.
	 We consider cases upon types of clusters $U$ and $V$.
    Note that this is impossible that either $U$ or $V$ is of type $C_3$, as otherwise we merge components of size $3$, and no component-respecting partition exists, a contradiction.
    If either $U$ or $V$ is of type $C_1$, then this cluster can fit the merged component, and the rebalance is local within $U$ and $V$, and costs at most $3\alpha$.
    \maciek{at most $2\alpha$ is sufficient}

    Finally, we consider the case where both $U$ and $V$ are of type $C_2$. Note that $(u,v)$ cannot both belong to a component of size $2$, as this would mean that \TAlg has the component of size $4$, a contradiction with the case assumption that the component-respecting partition exists. 
    Otherwise, if either $u$ or $v$ belongs to a component of size $2$, then it suffices to exchange components of size $1$ between $U$ and $V$.
    Finally, if $u$ and $v$ belong to components of size $1$, then we must place them in a cluster different from $U$ and $V$.
    Note that in such case, a $C_1$-type cluster $W$ exists, as otherwise no component-respecting partition exists. In this case \TAlg exchanges the nodes $u$ and $v$ with any two nodes of $W$, for the total cost $2\alpha$.
    \maciek{1 exchange costs $\alpha$}
\end{proof}

@@

A \emph{perfect partition} is a partition of nodes into $\ell$ clusters
such that it incurs no inter-cluster communication throughout the input sequence.
Equivalently, a perfect partition it an $\ell$-way cut of the communication graph of the input sequence.
In this section,
we give an algorithm for a restricted variant of  \OBRP{},
%\maciek{This is not sufficient! We also disallow serving requests remotely.}
where an optimal offline algorithm ($\OPT$) moves to this perfect partition
at the beginning and stays there perpetually.
%\maciek{Why do we have to assume that? This is just what OPT does, i.e., we can prove that this strategy (a) lower bounds the OPT's cost and (b) is sufficent for our analysis. But now is sounds like we restrict the model even further.}
We refer this variant as \PPOBRP{}.
The task of an online algorithm for \PPOBRP{} is to recover (or learn) the perfect partition while not paying too much relative to \OPT.
%
%\maciek{This is not a valid description of the model. We need the assumption that collocation is needed in our model. Otherwise OPT would just serve some requests remotely, and our analysis would not hold.}
The first request between any two non-collocated nodes reveals an edge of the communication graph.
Our online algorithm collocates them immediately and never separates them since by assumption \OPT does the same.
Revealed edges, up to any point in the sequence,
induce some connected subgraphs and we refer to them as \emph{components}.
An algorithm that always  keeps nodes belonging to the same component collocated is \emph{component-respecting}.

@@

We assume \OPT begins with the initial configuration
$P_I = I_1, \dots, I_{\ell}$ and moves to the final partition
$P_F = F_1, \dots, F_{\ell}$.
 The \emph{distance} of a configuration $P = C_1, \dots, C_{\ell}$ from the initial configuration is the number of nodes in $P$ that do not reside in their initial cluster.
    That is,
    $\dist(P, P_I) := \sum_{j=1}^{\ell} | C_j \setminus I_j |$. 
In other words,
at least $\dist(P, P_I)/2$ node swaps are required in order to reach the configuration $P$ from $P_I$, and thus
$\OPT \geq \Delta:= \dist(P_F, P_I) $.
 With each re-partitioning,
  \PPL moves to a configuration that minimizes the distance to the initial configuration $P_I$.
As a result,
\PPL never ends up in a configuration that is more than $\Delta$ (single-node) migrations away from $P_I$.
This invariant ensures that \PPL does not pay too much while recovering $P_F$.

>>>>

  In the online graph re-partitioning problem, we are given a sequence of pairwise communication requests between n nodes, revealed online.
  The objective is to service these requests by partitioning the nodes into $\ell$ clusters, each of size $k$, such that frequently communicating nodes are located in the same cluster.
  The partitioning can be updated by migrating nodes between clusters for a fixed cost.
  The goal is to jointly minimize the amount of inter-cluster communication and migration cost.

  So far, the best known lower bound for the competitive ratio of
	deterministic algorithms was $\Omega(k)$,
 by reduction from online paging.
  We improve this result significantly by providing the lower bound of $\Omega(k\cdot \ell)$.
  We follow up with an optimal algorithm for $k=3$.
  Finally,
  we provide a $O(k\cdot \ell)$-competitive algorithm for the variant
  where the optimal configuration is always a fixed partitioning (a.k.a~a \emph{configuration})
  that incurs zero communication cost through the whole request sequence.
\end{abstract}

@@

Main technical contribution of this paper is the lower bound $\Omega(k\cdot\ell)$ for the competitive ratio of any deterministic algorithm for online balanced re-partition problem (see Section~\ref{sec:lowerbound}).
The best known lower bound so far was $\Omega(k)$ \cite{repartition-disc}
which involves only two clusters.

@@

In Section~\ref{sec:ppl}, we study a restricted model of balanced re-partition, called perfect partition.
For that, we present a $O(k\cdot \ell)$-competitive algorithm, which is optimal.
We believe that this result is an important building block on the way to an optimal algorithm for the general model.

@@

The online \emph{balanced re-partitioning} problem (\OBRP{}) is defined as
follows. We are given a set of $n$ nodes,
initially arbitrarily partitioned into $\ell$~clusters,
each of size~$k$.
The input consists of a sequence of pairwise communication requests
$\sigma = (u_1,v_1),$ $(u_2,v_2),$ $(u_3,v_3), \ldots$,
where a pair $(u_t,v_t)$ indicates that nodes $u_t$ and $v_t$ exchange a fixed amount of data.
Nodes in $C \subset V$ are \emph{collocated}
if they reside in the same cluster.
An algorithm serves a communication request between two nodes
either \emph{locally} at cost~0
if they are collocated,
or \emph{remotely} at cost~1
if they are located in different clusters.
We refer to the two types of requests as \emph{internal}
and \emph{external} requests, respectively.
Before serving a request,
an online algorithm may \emph{re-partition},
%(i.e., \emph{reconfigure}).
i.e.,
it may (optionally) move some nodes into clusters different than their current clusters while respecting the capacity of every cluster. 
Only after a re-partitioning is completed,
the algorithm serves the  request.
The cost of migrating a node from one cluster to another
is~$\alpha \in \mathbb{Z}^+$.
For any algorithm $ALG$,
its cost,
denoted by $ALG(\sigma)$,
is the total cost of communications and
the cost of migrations performed by $ALG$ while serving the sequence $\sigma$.

@@

%We define the algorithm \TAlg in the following way.
Now we describe the algorithm \TAlg.
It partitions nodes into components, and
initially each node belongs to its own singleton component.
For each pair of nodes, \TAlg maintains a counter, initiated to $0$. 
Upon receiving a request to a pair that is not collocated at one cluster, it increases their counter by $1$.
If the counter for a pair $(u,v)$ reaches $\alpha$, \TAlg merges the components of $u$ and $v$, and moves to the closest component-respecting partitioning.
If no such partitioning exists, \TAlg resets all components to singletons, resets all counters to $0$, and ends the phase.
\maciek{Serving remotely: say explicitly}

@@

\begin{proof}
  Observe that when the re-partition is triggered by \TAlg, the resulting partitioning is component-respecting.
  Otherwise, if it does not exist, \TAlg simply ends the phase and performs no re-partition.

@@

    Consider a request between $u$ and $v$ that triggered the re-partition and let $U$ and $V$ be their respective clusters.
    Note that $U\neq V$,
	 as otherwise the request would not trigger a~re-partitioning.
	 We consider cases upon types of clusters $U$ and $V$.
    Note that this is impossible that either $U$ or $V$ is of type $C_3$, as otherwise we merge components of size $3$, and no component-respecting partitioning exists, a contradiction.
    If either $U$ or $V$ is of type $C_1$, then this cluster can fit the merged component, and the rebalance is local within $U$ and $V$, and costs at most $3\alpha$.
    \maciek{at most $2\alpha$ is sufficient}

    Finally, we consider the case where both $U$ and $V$ are of type $C_2$. Note that $(u,v)$ cannot both belong to a component of size $2$, as this would mean that \TAlg has the component of size $4$, a contradiction with the case assumption that the component-respecting partitioning exists. 
    Otherwise, if either $u$ or $v$ belongs to a component of size $2$, then it suffices to exchange components of size $1$ between $U$ and $V$.
    Finally, if $u$ and $v$ belong to components of size $1$, then we must place them in a cluster different from $U$ and $V$.
    Note that in such case, a $C_1$-type cluster $W$ exists, as otherwise no component-respecting partitioning exists. In this case \TAlg exchanges the nodes $u$ and $v$ with any two nodes of $W$, for the total cost $2\alpha$.
    \maciek{1 exchange costs $\alpha$}
\end{proof}

@@

A \emph{perfect partition} is a partitioning of nodes into $\ell$ clusters
such that it incurs no inter-cluster communication throughout the input sequence.
Equivalently, a perfect partitioning it an $\ell$-way cut of the communication graph of the input sequence.
In this section,
we give an algorithm for a restricted variant of  \OBRP{},
%\maciek{This is not sufficient! We also disallow serving requests remotely.}
where an optimal offline algorithm ($\OPT$) moves to this perfect partition
at the beginning and stays there perpetually.
%\maciek{Why do we have to assume that? This is just what OPT does, i.e., we can prove that this strategy (a) lower bounds the OPT's cost and (b) is sufficent for our analysis. But now is sounds like we restrict the model even further.}
We refer this variant as \PPOBRP{}.
The task of an online algorithm for \PPOBRP{} is to recover (or learn) the perfect partition while not paying too much relative to \OPT.
%
%\maciek{This is not a valid description of the model. We need the assumption that collocation is needed in our model. Otherwise OPT would just serve some requests remotely, and our analysis would not hold.}
The first request between any two non-collocated nodes reveals an edge of the communication graph.
Our online algorithm collocates them immediately and never separates them since by assumption \OPT does the same.
Revealed edges, up to any point in the sequence,
induce some connected subgraphs and we refer to them as \emph{components}.
An algorithm that always  keeps nodes belonging to the same component collocated is \emph{component-respecting}.

@@

We assume \OPT begins with the initial configuration
$P_I = I_1, \dots, I_{\ell}$ and moves to the final partitioning
$P_F = F_1, \dots, F_{\ell}$.
 The \emph{distance} of a configuration $P = C_1, \dots, C_{\ell}$ from the initial configuration is the number of nodes in $P$ that do not reside in their initial cluster.
    That is,
    $\dist(P, P_I) := \sum_{j=1}^{\ell} | C_j \setminus I_j |$. 
In other words,
at least $\dist(P, P_I)/2$ node swaps are required in order to reach the configuration $P$ from $P_I$, and thus
$\OPT \geq \Delta:= \dist(P_F, P_I) $.
 With each re-partitioning,
  \PPL moves to a configuration that minimizes the distance to the initial configuration $P_I$.
As a result,
\PPL never ends up in a configuration that is more than $\Delta$ (single-node) migrations away from $P_I$.
This invariant ensures that \PPL does not pay too much while recovering $P_F$.
